Unveiling the Optimal Approach for Neural Network-Based Cosmological Parameter Inference through Likelihood Implicit Methods

### 2023-06-25 paper/paper.tex
25/06

\newcommand{\justine}[1]{{\color{cyan}JZ: #1}}
\newcommand{\denise}[1]{{\color{red}DL: #1}}

@@

   \maketitle
%
%-------------------------------------------------------------------

\section{Introduction}
\section{Theory Background}
\subsection{Weak lensing formalism}
\subsection{Lognormal Modeling}
For several cosmological purposes, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model has the advantage of offering the rapid generation of the matter or the convergence field while allowing the extraction of information beyond the Gaussian one. 
Although several studies demonstrated that this model fails in describing the 3D field, it properly describes the 2D convergence field.

@@

Having defined the survey to reproduce in terms of galaxy number density, redshifts, and shape noise, we can sample the Lognormal random map $\kappa$.  using \autoref{Eq:log_norm_kappa} and \autoref{Eq:log_norm_corr}.  First, we need to compute the theoretical auto $C_{\ell}^{i}$ and cross-correlated $C_{\ell}^{ij}$ angular power spectrum for each tomographic bin. These theoretical predictions are computed using the public library  \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-mosmo}}. Then the 1-D $C_{\ell}$ are projected into 2-D grids of the same size as the desired final convergence maps and the Gaussian covariances $\xi^{ij}_y(\theta)$ are computed through \autoref{Eq:log_norm_corr}. To take into account the cross-correlation between different redshift bins, we first stack together the vector $\xi^{ij}_y(\theta)$ to build a covariance matrix having the auto-correlation on the main diagonal and the cross-correlation off-diagonal. Then we perform an eigenvalues decomposition. This decomposition enables us to decorrelate the maps

\begin{equation}
    \bm{\Sigma}= 
    \begin{pmatrix}
    C_{\ell}^{11} & C_{\ell}^{12} & \cdots & C_{\ell}^{1n} \\
    C_{\ell}^{21} & C_{\ell}^{22} & \cdots & C_{\ell}^{2n} \\
    \vdots  & \vdots  & \ddots & \vdots  \\
    C_{\ell}^{n1} & C_{\ell}^{n2} & \cdots & C_{\ell}^{nn} 
    \end{pmatrix}
\end{equation}
%--------------------------------------------------------------------
\section{Simulations}
\subsection{Data generation}
Our analysis is based on a standard flat $\Lambda$CDM cosmological model, with the following parameters: the baryonic density fraction $\Omega_b$, the total matter density fraction $\Omega_m$, the Hubble parameter $h_0$, the spectral index $n_s$, the amplitude of the primordial power spectrum $\sigma_8$ and the dark energy parameter $w_0$. The priors used in the simulations and in inference process are listed in \autoref{tab:prior}, according to \citet{zhang2022transitioning}.
The simulation process proceeds as follows. Given the cosmological parameters, we compute the theoretical power- and cross- spectra using the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}} \citep{Campagne_2023}. To calculate the 
lognormal shift parameter, we adopt the \texttt{Cosmomentum} code \citep{friedrich2018density, friedrich2020primordial}, which utilizes perturbation theory to compute the cosmology-dependent shift parameters. It should be noted that the calculation of the shift parameters is based on a window function, while our pixels are rectangular. Following \citet{boruah2022map}, we compute the shift parameters at a characteristic scale, $R=\Delta L/\pi$  with $\Delta L$, the resolution of the pixels.  Moreover, as pointed out by \citet{boruah2022map}, this perturbation theory-based approach may not provide accurate results at small scales. However, since the main objective of this paper is to compare different compression strategies, we are not concerned about the potential implications of this approximation. In future applications of the presented inference pipeline, the simulation procedure will be based on N-body simulations.

@@

\begin{table}
	\centering
	\caption{ Prior and fiducial values used for the analyses. 
 The symbol $\mathcal{N}_T$ represents a Truncated Normal distribution. The lower bound of the support for the $\Omega_c$ distribution is set to zero, while the lower and upper bounds for the $w_0$ distribution are set to -2.0 and -0.3, respectively.}
	\begin{tabular}{lcc} 
		\hline \hline
		Parameter  & Prior & Fiducial value \\
		$\Omega_c$ & $\mathcal{N}_T$ (0.2664, 0.2) & 0.2664 \\
		$\Omega_b$ & $\mathcal{N}$ (0.0492, 0.006) & 0.0492 \\
		$\sigma_8$ & $\mathcal{N}$ (0.831, 0.14) & 0.831 \\
		$h$ & $\mathcal{N}$ (0.6727, 0.063) & 0.6727\\
		$n_s$ & $\mathcal{N}$ (0.9645, 0.08) & 0.9645 \\
		$w_{0}$ &  $\mathcal{N}_T$ (-1.0, 0.9) &  -1.0 \\
		\hline
	\end{tabular}
	\label{tab:prior}
\end{table}
%--------------------------------------------------------------------

@@

@@

@@

@@

@@

@@

@@

Ultimately, for a given simulation model, the two approaches should converge to the same posterior.
%--------------------------------------------------------------------
\subsection{Explicit Inference}
\subsubsection{Power spectrum}
Although this is not a full-field approach, we aim to compare the latest to a more standard approach. 
To obtain the probability distribution of the cosmological parameters using a summary-statistics-methods, we focus on the 2-point statistics, specifically, on the angular power spectra $C_{\ell}$.
The correlation between two generic tomographic bins $i$ and $j$ can be expressed as:
\begin{equation}
    \left \langle \kappa^{i}(\ell)\kappa^{j*}(\ell)\right \rangle =(2\pi)^2\delta_D(\ell-\ell')C_{\ell}^{ij},
\end{equation}
We assume a Gaussian likelihood with a cosmology-independent covariance matrix:
\begin{equation}
    \log{\mathcal{L}}(\bm{\theta})=-\frac{1}{2}[\bm{d}-\bm{\mu}(\bm{\theta})]^{T}\bm{C}^{-1}[\bm{d}-\bm{\mu}(\bm{\theta})].
\end{equation}
To compute the expected theoretical predictions $\bm{\mu}(\bm{\theta})$ we use the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}}. 
 The Covariance matrix $\bm{C}$ of the observables is computed at the fiducial cosmology, shown in \autoref{tab:prior}, using the same theoretical library. In {\texttt{jax-cosmo}}, the Gaussian covariance matrix is defined as:
\begin{equation}
    \text{Cov}(C_{\ell},C_{\ell}')=\frac{1}{f_{sky}(2 \ell+1)}\left(C_{\ell}+\frac{\sigma_{\epsilon}^2}{2n_s}\right)
\end{equation}
where $f_{sky}$ is the fraction of sky observed by the survey, and $n_s$ is the number density of galaxies. 
To obtain the data vector $\bm{d}$, containing the auto- and the cross-power spectra for each tomographic bin, we use the \href{https://lenstools.readthedocs.io/en/latest/lenstool} {\texttt{LensTools}} package \citep{2016A&C....17...73P} on a single noisy simulated map. 
\subsubsection{Full field and HMC}
The following steps outline our methods. We draw cosmological parameters from the priors illustrated in \autoref{tab:prior}. Given values for the cosmological parameters, we generate the log-normal field as described in the previous sections. So, in other words, the simulator becomes the physical model generating a non-linear representation of the convergence map. The measure of the convergence for each pixel and for each bin predicted from the physical model will differ from the ones from real observations due to the noise.  This is taken into account in the likelihood. Specifically, for LSST Y10 the number of galaxies for each pixel should be sufficiently high that for the central limit theorem, we can assume that the observation is characterized by a Gaussian noise $\sigma_n^2=\sigma_e^2/N_s$, with $N_s$ the total number of source galaxies per bin and pixel. Given $\sigma_n^2$ the variance of this Gaussian likelihood, its form in the log form can be expressed as:
\begin{equation}
    \log{\mathcal{L}}(\bm{\theta})=
    \sum_i^{N_{pix}} \sum_{j}^{N_{bins}} \log{P(\kappa^{obs}_{i,j}|\kappa_{i,j},\bm{\theta})}
    =\sum_i^{N_{pix}} \sum_{j}^{N_{bins}}\frac{[\kappa_{i,j}-\kappa^{obs}_{i,j}]^2}{2\sigma_n^2}
\end{equation}
Not relying on any summary statistics, the full-field approach typically results in a high-dimensional problem, requiring more sophisticated statistical techniques. To sample the posterior distribution for $\bm{\theta}$, we use a Hamiltonian Monte Carlo (HMC) algorithm. The HMC algorithm is particularly helpful in high-dimensional spaces where a large number of steps are required to effectively explore the space. It improves the sampling process by using the information contained in the gradients to guide the sampling process. Since the code is implemented in Jax, the gradients are accessible via automatic differentiation.

%--------------------------------------------------------------------
\subsection{Implicit Inference}

\subsubsection{Benchmark compression scheme}
There are several approaches available for training the compressor, and this section aims to provide an overview of these different methods used in previous works. 
In \autoref{Sec:results}, we will compare the outcomes of these various strategies. Specifically, using the same prior $p(\bm {\theta})$ for the cosmological parameter $\bm{\theta}$ and the simulator described in \autoref{Sec:results}, which samples $\bm {x}$ from $p(\bm{x}| \bm{\theta})$ and produces a set of mock observations $\bm{x_0}$, we will compare the posterior distribution $p(\bm{\theta}|\bm{x})$ obtained form each inference pipeline. This analysis will help to determine the optimal compression strategy, enabling us to extract the maximum amount of information possible given the fixed cosmological parameters.

@@

A third possible scenario consists in minimizing the Negative Log Likelihood. First of all, let us stating that minimizing the negative log likelihood is equivalent to the Maximum Likelihood Estimation (MLE). The proof of this statement is straightforward. As suggested by the name, this optimization problem involves maximizing the likelihood distribution $p(\bm{x}|\bm{\theta})$
    \bm{\theta}_{ML}= \operatorname*{argmax}_{\bm{\theta}}p(\bm{x}|\bm{\theta}).
where
    p(\bm{x}|\bm{\theta})= \prod_{i-1}^n p(x_i|\bm{\theta}).
Applying the log function in this context is done for several reasons, e.g. reducing the potential of underflow when the likelihoods are small or allowing the conversion of a product into a summation. 
Then, remembering that $-\log$ is strictly decreasing, the following are equivalent:
    \operatorname*{argmax}_{\bm{\theta}}\sum_{i-1}^n \log{p(x_i|\bm{\theta})}= \operatorname*{argmin}_{\bm{\theta}}-\sum_{i-1}^n \log{p(x_i|\bm{\theta})}.

\paragraph{\textcolor{violet}{Information Maximising Neural Networks (IMNNs)}}
%--------------------------------------------------------------------
\paragraph{\textcolor{violet}{Variational Mutual Information Maximization (VMIM)}}
The Variational Mutual Information Maximization (VMIM) has been used for cosmological inference problems for the first time by \citet{jeffrey2021likelihood}. This approach aims to maximize the mutual information $I(\bm{t}, \bm {\theta})$ between the cosmological parameters $\bm{\theta}$ and the summary statistics $\bm t$ by training a neural network with parameters $\phi$. 
Mathematically, it is defined as follows:
\begin{align}\label{Eq:mutual_information}
    I(\bm{t}, \bm {\theta}) &= D_{KL}(p(\bm {t}, \bm {\theta})||p(\bm {t})p(\bm {\theta})) \\ \nonumber
    &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm \theta)\log{\left( \frac{ p(\bm {t}, \bm {\theta})}{ p(\bm {t}) p(\bm {\theta})} \right)} \\ \nonumber
    &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm {\theta})\log{\left( \frac{ p(\bm {\theta} | \bm {t} )}{ p(\bm {\theta})} \right)} \\ \nonumber
        &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm {\theta})\log{p(\bm {\theta} | \bm {t} )} - \int d^n \bm{\theta}  d^n \bm{t} p(\bm t, \bm {\theta})\log{p(\bm {\theta})} \\ \nonumber
    &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm {\theta})\log{p(\bm {\theta} | \bm {t} )} - \int d^n \bm{\theta} p(\bm {\theta})\log{p(\bm {\theta})} \\ \nonumber
    &= \mathbb{E}_{p(\bm {t}, \bm {\theta})} [\log{p(\bm {\theta} | \bm {t} )}]- \mathbb{E}_{p(\bm {\theta})} [\log{p(\bm {\theta})}] \\ \nonumber
    &= \mathbb{E}_{p(\bm {t}, \bm {\theta})} [\log{p(\bm {\theta} | \bm {t} )}]- H(\bm {\theta});
\end{align}

where $D_{KL}$ is the Kullback-Leibler divergence \citep{kullback1951information}, $p(\bm {t}, \bm {\theta})$ is the joint probability distribution of summary statistics and cosmological parameters, and $H(\bm {\theta})$ represents the \textit{entropy} of the distribution of cosmological parameters.  
In other words, the mutual information measures the amount of information contained in the summary statistics $\bm t$ about the cosmological parameters $\bm \theta$.
Given the original high-dimensional data vector $\bm d$, the information $\bm t$ is extracted using $\bm {t}=F_{\bm {\phi}}(\bm {d})$. 
The goal is to find the parameters $\bm {\phi}$ that maximize the mutual information between the summary and cosmological parameters:
\begin{equation}
   \bm {\phi}^*= \operatorname*{argmax}_{\bm {\phi}} I(F_{\bm {\phi}}(\bm{d}), \bm {\theta}).
\end{equation}
However, the mutual information expressed in \autoref{Eq:mutual_information} is not tractable. To overcome this limitation,  several approaches have been developed, relying on tractable bounds that allow the training of a deep neural network to optimize the mutual information. In this work, we adopt the same strategy used by \citet{jeffrey2021likelihood} and use the variational lower bound \citep{barber2003information}:
\begin{equation}\label{Eq:variational_lower_bound}
    I(\bm{t}, \bm{\theta}) \ge \mathbb{E}_{p(\bm {t}, \bm {\theta})} [\log{q(\bm {\theta} |\bm{t} ; \bm{\phi}')}]- H(\bm {\theta}).
\end{equation}
Here, the variational conditional distribution $\log{q(\bm {\theta} |\bm{t} ; \bm{\phi}')}$ is introduced to approximate the true posterior distribution $p(\bm{\theta}|\bm {t})$. 
Since the entropy of the cosmological parameters is constant for a fixed training set, the optimization problem based on the lower bound in  \autoref{Eq:variational_lower_bound} can be written as:
\begin{equation}
    \operatorname*{argmax}_{\bm {\phi}, \bm {\phi}'}\mathbb{E}_{p(\bm {d}, \bm {\theta})} [\log{q(\bm {\theta} |F_{\bm {\phi}}(\bm {d}) ; \bm{\phi}')}].
\end{equation}

\subsubsection{Inference strategy}
%--------------------------------------------------------------------
To evaluate compression strategies, we propose a comparison of posteriors $p(\theta|s(x))$, where $s(x)$ is the summary statistic. Indeed, the more informative the summary statistics, the closer the posterior $p(\theta | s(x))$ is to the true posterior $p (\theta | x)$. To this end, we propose to do Neural Posterior Estimation (NPE) that aims to directly learn from simulations the posterior distribution

\begin{equation}
    p(\theta | x) \propto \int p(x | \theta, z) p(z|\theta) p(\theta) dz\,.
\end{equation}

To achieve this, we train a  conditional Normalizing Flow (NF) $p_{\varphi} (\theta | s(x)) $ by minimizing the negative log-probability in $(\theta, s(x))$ space:

\begin{equation}
    \mathcal{L}_{\rm NLL} = \mathbb{E}_{p(\theta, s(x))} \left[ - \log p_{\varphi} (\theta | s(x))  \right],
    \label{eq:nll}
\end{equation}

@@

% talk about evaluation of approx posterior at a given obs

>>>>

\newcommand{\justine}[1]{{\color{cyan}JZ: #1}}
\newcommand{\denise}[1]{{\color{red}DL: #1}}
\newcommand{\FL}[1]{{\color{magenta}FL: #1}}

@@

   \maketitle

%-------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{Introduction}
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
Ultimately, for a given simulation model, the two approaches should converge to the same posterior.
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------
\section{The SBILens framework}
\subsection{Lognormal Modeling}
For several cosmological purposes, the non-Gaussian field can be modeled as a Lognormal field \citep{coles1991lognormal,bohm2017bayesian}.
This model has the advantage of offering the rapid generation of the matter or the convergence field while allowing the extraction of information beyond the Gaussian one. 
Although several studies demonstrated that this model fails in describing the 3D field, it properly describes the 2D convergence field.

@@

Having defined the survey to reproduce in terms of galaxy number density, redshifts, and shape noise, we can sample the Lognormal random map $\kappa$.  using \autoref{Eq:log_norm_kappa} and \autoref{Eq:log_norm_corr}.  First, we need to compute the theoretical auto $C_{\ell}^{i}$ and cross-correlated $C_{\ell}^{ij}$ angular power spectrum for each tomographic bin. These theoretical predictions are computed using the public library  \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}}. Then the 1-D $C_{\ell}$ are projected into 2-D grids of the same size as the desired final convergence maps and the Gaussian covariances $\xi^{ij}_y(\theta)$ are computed through \autoref{Eq:log_norm_corr}. To take into account the cross-correlation between different redshift bins, we first stack together the vector $\xi^{ij}_y(\theta)$ to build a covariance matrix having the auto-correlation on the main diagonal and the cross-correlation off-diagonal. Then we perform an eigenvalues decomposition. This decomposition enables us to decorrelate the maps

\begin{equation}
    \bm{\Sigma}= 
    \begin{pmatrix}
    C_{\ell}^{11} & C_{\ell}^{12} & \cdots & C_{\ell}^{1n} \\
    C_{\ell}^{21} & C_{\ell}^{22} & \cdots & C_{\ell}^{2n} \\
    \vdots  & \vdots  & \ddots & \vdots  \\
    C_{\ell}^{n1} & C_{\ell}^{n2} & \cdots & C_{\ell}^{nn} 
    \end{pmatrix}
\end{equation}
%------------------------------------------------------------------
%------------------------------------------------------------------
%--------------------------------------------------------------------
\subsection{Data generation}
Our analysis is based on a standard flat $\Lambda$CDM cosmological model, with the following parameters: the baryonic density fraction $\Omega_b$, the total matter density fraction $\Omega_m$, the Hubble parameter $h_0$, the spectral index $n_s$, the amplitude of the primordial power spectrum $\sigma_8$ and the dark energy parameter $w_0$. The priors used in the simulations and in inference process are listed in \autoref{tab:prior}, according to \citet{zhang2022transitioning}.
The simulation process proceeds as follows. Given the cosmological parameters, we compute the theoretical power- and cross- spectra using the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}} \citep{Campagne_2023}. To calculate the 
lognormal shift parameter, we adopt the \texttt{Cosmomentum} code \citep{friedrich2018density, friedrich2020primordial}, which utilizes perturbation theory to compute the cosmology-dependent shift parameters. It should be noted that the calculation of the shift parameters is based on a window function, while our pixels are rectangular. Following \citet{boruah2022map}, we compute the shift parameters at a characteristic scale, $R=\Delta L/\pi$  with $\Delta L$, the resolution of the pixels.  Moreover, as pointed out by \citet{boruah2022map}, this perturbation theory-based approach may not provide accurate results at small scales. However, since the main objective of this paper is to compare different compression strategies, we are not concerned about the potential implications of this approximation. In future applications of the presented inference pipeline, the simulation procedure will be based on N-body simulations.

@@

	\caption{ Prior and fiducial values used for the analyses. 
 The symbol $\mathcal{N}_T$ represents a Truncated Normal distribution. The lower bound of the support for the $\Omega_c$ distribution is set to zero, while the lower and upper bounds for the $w_0$ distribution are set to -2.0 and -0.3, respectively.}
	\begin{tabular}{lcc} 
		Parameter  & Prior & Fiducial value \\
		$\Omega_c$ & $\mathcal{N}_T$ (0.2664, 0.2) & 0.2664 \\
		$\Omega_b$ & $\mathcal{N}$ (0.0492, 0.006) & 0.0492 \\
		$\sigma_8$ & $\mathcal{N}$ (0.831, 0.14) & 0.831 \\
		$h$ & $\mathcal{N}$ (0.6727, 0.063) & 0.6727\\
		$n_s$ & $\mathcal{N}$ (0.9645, 0.08) & 0.9645 \\
		$w_{0}$ &  $\mathcal{N}_T$ (-1.0, 0.9) &  -1.0 \\
	\label{tab:prior}
%------------------------------------------------------------------
%------------------------------------------------------------------
%------------------------------------------------------------------

@@

%------------------------------------------------------------------
\section{Experiment}
\subsection{Explicit Inference}
\subsubsection{Power spectrum}
Although this is not a full-field approach, we aim to compare the latest to a more standard approach. 
To obtain the probability distribution of the cosmological parameters using a summary-statistics-methods, we focus on the 2-point statistics, specifically, on the angular power spectra $C_{\ell}$.
The correlation between two generic tomographic bins $i$ and $j$ can be expressed as:

@@

    \left \langle \kappa^{i}(\ell)\kappa^{j*}(\ell)\right \rangle =(2\pi)^2\delta_D(\ell-\ell')C_{\ell}^{ij},

@@

We assume a Gaussian likelihood with a cosmology-independent covariance matrix:

@@

    \log{\mathcal{L}}(\bm{\theta})=-\frac{1}{2}[\bm{d}-\bm{\mu}(\bm{\theta})]^{T}\bm{C}^{-1}[\bm{d}-\bm{\mu}(\bm{\theta})].

@@

To compute the expected theoretical predictions $\bm{\mu}(\bm{\theta})$ we use the public library \href{https://github.com/DifferentiableUniverseInitiative/jax_cosmo}{\texttt{jax-cosmo}}. 
 The Covariance matrix $\bm{C}$ of the observables is computed at the fiducial cosmology, shown in \autoref{tab:prior}, using the same theoretical library. In {\texttt{jax-cosmo}}, the Gaussian covariance matrix is defined as:

@@

    \text{Cov}(C_{\ell},C_{\ell}')=\frac{1}{f_{sky}(2 \ell+1)}\left(C_{\ell}+\frac{\sigma_{\epsilon}^2}{2n_s}\right)

@@

where $f_{sky}$ is the fraction of sky observed by the survey, and $n_s$ is the number density of galaxies. 
To obtain the data vector $\bm{d}$, containing the auto- and the cross-power spectra for each tomographic bin, we use the \href{https://lenstools.readthedocs.io/en/latest/lenstool} {\texttt{LensTools}} package \citep{2016A&C....17...73P} on a single noisy simulated map. 
\subsubsection{Full field and HMC}
The following steps outline our methods. We draw cosmological parameters from the priors illustrated in \autoref{tab:prior}. Given values for the cosmological parameters, we generate the log-normal field as described in the previous sections. So, in other words, the simulator becomes the physical model generating a non-linear representation of the convergence map. The measure of the convergence for each pixel and for each bin predicted from the physical model will differ from the ones from real observations due to the noise.  This is taken into account in the likelihood. Specifically, for LSST Y10 the number of galaxies for each pixel should be sufficiently high that for the central limit theorem, we can assume that the observation is characterized by a Gaussian noise $\sigma_n^2=\sigma_e^2/N_s$, with $N_s$ the total number of source galaxies per bin and pixel. Given $\sigma_n^2$ the variance of this Gaussian likelihood, its form in the log form can be expressed as:
    \log{\mathcal{L}}(\bm{\theta})=
    \sum_i^{N_{pix}} \sum_{j}^{N_{bins}} \log{P(\kappa^{obs}_{i,j}|\kappa_{i,j},\bm{\theta})}
    =\sum_i^{N_{pix}} \sum_{j}^{N_{bins}}\frac{[\kappa_{i,j}-\kappa^{obs}_{i,j}]^2}{2\sigma_n^2}
Not relying on any summary statistics, the full-field approach typically results in a high-dimensional problem, requiring more sophisticated statistical techniques. To sample the posterior distribution for $\bm{\theta}$, we use a Hamiltonian Monte Carlo (HMC) algorithm. The HMC algorithm is particularly helpful in high-dimensional spaces where a large number of steps are required to effectively explore the space. It improves the sampling process by using the information contained in the gradients to guide the sampling process. Since the code is implemented in Jax, the gradients are accessible via automatic differentiation. 
\subsection{Implicit Inference}
\subsubsection{Need for data compression}
\subsubsection{Benchmark compression scheme}
There are several approaches available for training the compressor, and this section aims to provide an overview of these different methods used in previous works. 
In \autoref{Sec:results}, we will compare the outcomes of these various strategies. Specifically, using the same prior $p(\bm {\theta})$ for the cosmological parameter $\bm{\theta}$ and the simulator described in \autoref{Sec:results}, which samples $\bm {x}$ from $p(\bm{x}| \bm{\theta})$ and produces a set of mock observations $\bm{x_0}$, we will compare the posterior distribution $p(\bm{\theta}|\bm{x})$ obtained form each inference pipeline. This analysis will help to determine the optimal compression strategy, enabling us to extract the maximum amount of information possible given the fixed cosmological parameters.

@@

\FL{I think we should have this section after the VMIM}

\FL{Recognizing that the uncertainty on different cosmological parameters will vary, a third
class of inverse variance weighted MSE was proposed in \cite{fluri2018cosmological} with the idea of
ensuring that each parameter contributes fairly to the overall loss by taking into account its uncertainty. The loss function typically takes the following form:
\begin{equation}
    \mathcal{L} = \frac{1}{2} \log(|\Sigma|) + \frac{1}{2}(t - \theta)^t \Sigma^{-1} (t - \theta) 
\end{equation}
where $t$ is the summary statistics and $\Sigma$ is the covariance matrix of representing the uncertainty on the cosmological parameters $\theta$. Both $t$ and $\Sigma$ can be outputs of the compression network, i.e. $f_\theta(x)=(t, \Sigma)$.

One recognizes here the expression of a Gaussian probability function, and this expression can thus be straightforwardly related to the VMIM case by simply assuming a Gaussian distribution as the variational approximation for the posterior $q(\theta | x) = \mathcal{N}(\theta; t, \Sigma)$.

This means that this loss function is actually sub-optimal for two reasons:
\begin{itemize}
    \item The summary extracted by the neural network $t$ is, similarly to the MSE case only an estimate of the mean of 
    the posterior distribution, which is not guaranteed to be sufficient.
    \item Because of the Gaussian variational assumption, the variational posterior can potentially be biased with respect to the true posterior, and thus the mean of this Gaussian approximation may be biased with respect to the true posterior mean.
\end{itemize}
 A simple MSE would thus seem to be preferable.
}
% A third possible scenario consists in minimizing the Negative Log Likelihood. First of all, let us stating that minimizing the negative log likelihood is equivalent to the Maximum Likelihood Estimation (MLE). The proof of this statement is straightforward. As suggested by the name, this optimization problem involves maximizing the likelihood distribution $p(\bm{x}|\bm{\theta})$
% \begin{equation}
%     \bm{\theta}_{ML}= \operatorname*{argmax}_{\bm{\theta}}p(\bm{x}|\bm{\theta}).
% \end{equation}
% where
% \begin{equation}
%     p(\bm{x}|\bm{\theta})= \prod_{i-1}^n p(x_i|\bm{\theta}).
% \end{equation}
% Applying the log function in this context is done for several reasons, e.g. reducing the potential of underflow when the likelihoods are small or allowing the conversion of a product into a summation. 
% Then, remembering that $-\log$ is strictly decreasing, the following are equivalent:
% \begin{equation}
%     \operatorname*{argmax}_{\bm{\theta}}\sum_{i-1}^n \log{p(x_i|\bm{\theta})}= \operatorname*{argmin}_{\bm{\theta}}-\sum_{i-1}^n \log{p(x_i|\bm{\theta})}.
% \end{equation}

%--------------------------------------------------------------------
\paragraph{\textcolor{violet}{Information Maximising Neural Networks (IMNNs)}}
%--------------------------------------------------------------------
\paragraph{\textcolor{violet}{Variational Mutual Information Maximization (VMIM)}}
The Variational Mutual Information Maximization (VMIM) has been used for cosmological inference problems for the first time by \citet{jeffrey2021likelihood}. This approach aims to maximize the mutual information $I(\bm{t}, \bm {\theta})$ between the cosmological parameters $\bm{\theta}$ and the summary statistics $\bm t$ by training a neural network with parameters $\phi$. 
Mathematically, it is defined as follows:
\begin{align}\label{Eq:mutual_information}
    I(\bm{t}, \bm {\theta}) &= D_{KL}(p(\bm {t}, \bm {\theta})||p(\bm {t})p(\bm {\theta})) \\ \nonumber
    &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm \theta)\log{\left( \frac{ p(\bm {t}, \bm {\theta})}{ p(\bm {t}) p(\bm {\theta})} \right)} \\ \nonumber
    &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm {\theta})\log{\left( \frac{ p(\bm {\theta} | \bm {t} )}{ p(\bm {\theta})} \right)} \\ \nonumber
        &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm {\theta})\log{p(\bm {\theta} | \bm {t} )} - \int d^n \bm{\theta}  d^n \bm{t} p(\bm t, \bm {\theta})\log{p(\bm {\theta})} \\ \nonumber
    &= \int d^n \bm{\theta} d^n \bm{t} p(\bm t, \bm {\theta})\log{p(\bm {\theta} | \bm {t} )} - \int d^n \bm{\theta} p(\bm {\theta})\log{p(\bm {\theta})} \\ \nonumber
    &= \mathbb{E}_{p(\bm {t}, \bm {\theta})} [\log{p(\bm {\theta} | \bm {t} )}]- \mathbb{E}_{p(\bm {\theta})} [\log{p(\bm {\theta})}] \\ \nonumber
    &= \mathbb{E}_{p(\bm {t}, \bm {\theta})} [\log{p(\bm {\theta} | \bm {t} )}]- H(\bm {\theta});
\end{align}

where $D_{KL}$ is the Kullback-Leibler divergence \citep{kullback1951information}, $p(\bm {t}, \bm {\theta})$ is the joint probability distribution of summary statistics and cosmological parameters, and $H(\bm {\theta})$ represents the \textit{entropy} of the distribution of cosmological parameters.  
In other words, the mutual information measures the amount of information contained in the summary statistics $\bm t$ about the cosmological parameters $\bm \theta$.
Given the original high-dimensional data vector $\bm d$, the information $\bm t$ is extracted using $\bm {t}=F_{\bm {\phi}}(\bm {d})$. 
The goal is to find the parameters $\bm {\phi}$ that maximize the mutual information between the summary and cosmological parameters:
\begin{equation}
   \bm {\phi}^*= \operatorname*{argmax}_{\bm {\phi}} I(F_{\bm {\phi}}(\bm{d}), \bm {\theta}).
\end{equation}
However, the mutual information expressed in \autoref{Eq:mutual_information} is not tractable. To overcome this limitation,  several approaches have been developed, relying on tractable bounds that allow the training of a deep neural network to optimize the mutual information. In this work, we adopt the same strategy used by \citet{jeffrey2021likelihood} and use the variational lower bound \citep{barber2003information}:
\begin{equation}\label{Eq:variational_lower_bound}
    I(\bm{t}, \bm{\theta}) \ge \mathbb{E}_{p(\bm {t}, \bm {\theta})} [\log{q(\bm {\theta} |\bm{t} ; \bm{\phi}')}]- H(\bm {\theta}).
\end{equation}
Here, the variational conditional distribution $\log{q(\bm {\theta} |\bm{t} ; \bm{\phi}')}$ is introduced to approximate the true posterior distribution $p(\bm{\theta}|\bm {t})$. 
Since the entropy of the cosmological parameters is constant for a fixed training set, the optimization problem based on the lower bound in  \autoref{Eq:variational_lower_bound} can be written as:
\begin{equation}
    \operatorname*{argmax}_{\bm {\phi}, \bm {\phi}'}\mathbb{E}_{p(\bm {d}, \bm {\theta})} [\log{q(\bm {\theta} |F_{\bm {\phi}}(\bm {d}) ; \bm{\phi}')}].
\end{equation}
%--------------------------------------------------------------------
%--------------------------------------------------------------------
%--------------------------------------------------------------------
\subsubsection{Inference strategy}

To evaluate compression strategies, we propose a comparison of posteriors $p(\theta|s(x))$, where $s(x)$ is the summary statistic. Indeed, the more informative the summary statistics, the closer the posterior $p(\theta | s(x))$ is to the true posterior $p (\theta | x)$. To this end, we propose to do Neural Posterior Estimation (NPE) \justine{describe a bit more what it is, sbi method etc} that aims to directly learn from simulations the posterior distribution
\begin{equation}
    p(\theta | x) \propto \int p(x | \theta, z) p(z|\theta) p(\theta) dz\,.
\end{equation}
To achieve this, we train a  conditional Normalizing Flow (NF) \justine{as proposed by ... (REF) / neural simulation-based inference} $p_{\varphi} (\theta | s(x)) $ by minimizing the negative log-probability in $(\theta, s(x))$ space:
\begin{equation}
    \mathcal{L}_{\rm NLL} = \mathbb{E}_{p(\theta, s(x))} \left[ - \log p_{\varphi} (\theta | s(x))  \right].
    \label{eq:nll}
\end{equation}

\justine{talk about evaluation of approx posterior at a given obs}

@@
